---
title: "Processing and analysis of nanopore sequences"
author: "Marko Suokas"
format: pdf
pdf-engine: lualatex
editor: visual
mainfont: Aptos
always_allow_html: yes
header-includes:
   \usepackage[dvipsnames]{xcolor}
   \definecolor{teal}{rgb}{0.0, 0.5, 0.5}
   \definecolor{ivory}{rgb}{1.0, 1.0, 0.94}
---

```{r, include = F}

# This allows to use different font sizes inside code chunks
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

# Global options
knitr::opts_chunk$set(warning = F, message = F) 

# Necessary R libraries
library(ggsci)
library(mia)
library(scater)
library(miaViz)
library(vegan)
library(Biostrings)
library(ShortRead)
library(tidyverse)
library(kableExtra)
library(ggthemes)
library(patchwork)
```

#### Preprocess reads

Dorado does not support demultiplexing dual indexes located on both the 5' and 3' ends. Additionally, in ligated libraries, the reads can appear in either orientation. To address this, we use `cutadapt` for demultiplexing. Index pairs are identified using the linked adapters approach in both forward and reverse orientations, after which scripts are applied to reverse complement the reverse reads. Finally, the reads are merged.

#### Extract Forward Reads

You can extract forward reads using the following ion torrent adapter sequences with following command:

``` bash
cutadapt --trimmed-only -e 0.05 -o reanalysis/fwd.fastq.gz \
-g "CCATCTCATCCCTGCGTGTCTCCGACTCAG;o=30...ATCACCGACTGCCCATAGAGAGG;o=23" \
evseq_rebasecall.fastq.gz
```

#### Extract Reverse Reads

To extract reverse reads, use the reverse-complement of adapter sequences

``` bash
cutadapt --trimmed-only -e 0.05 -o reanalysis/reverse.fastq.gz \
-g "CCTCTCTATGGGCAGTCGGTGAT;o=23...CTGAGTCGGAGACACGCAGGGATGAGATGG;o=30" \
evseq_rebasecall.fastq.gz 
```

#### Reverse Complement Reverse Reads

``` bash
seqkit seq -rp --seq-type DNA -o revcomp.fastq.gz \
reverse.fastq.gz
```

#### Merge Forward and Reverse Reads

For the final step, you can merge forward and reverse reads with the same base name from two directories. Here's a simple bash command for that:

``` bash
cat forward.fastq.gz rev.fastq.gz >raw.fastq.gz
```

#### Read quality

Nanopore sequencing Phred scores should be interpreted with caution, as per-base accuracy cannot be directly determined from the electrical signal. Instead, scores are estimates based on the confidence of the basecalling model rather than a direct probability of sequencing error. As a result, traditional quality filtering approaches may not always be appropriate.

ONT instead calculates an estimated cumulative error rate and converts it into an average quality score per read. The following code computes this metric and generates a violin plot showing the distribution of average sequence quality per read for each sample.

In this set we have 132 samples. We will analyze quality of only 48 samples in order to keep total number in plot manageable.

```{r nanopore_quality, eval = F, size = "small"}
# Define path
source_dir <- "data/reads/"
# A function to calculate so called AvgQual
extract_nano_qscores <- function(file) {
    # Read fastq
    fq <- readFastq(file)
    # Extract q-values to a matrix
    qmat <- as(quality(fq), "matrix")
    # Convert values to error probabilities
    error_probs <- 10^(-qmat/10)
    # Compute number of expected errors per read
    total_errors <- rowSums(error_probs, na.rm = T)
    # Length per read
    read_lenghts <- rowSums(!is.na(qmat))
    # Compute "ONT-style" Q-score
    ont_qscores <- -10*log10(total_errors/read_lenghts)
    # Trim file names
    sampleid <- sub("_.*|\\.fastq\\.gz$", "", basename(file))
    # Create a dataframe
    qdata <- tibble(
        nanopore_qscore = ont_qscores,
        sampleid = sampleid)
    return(qdata)
}

# Create list of files
files <- list.files(source_dir, pattern = "\\.fastq\\.gz$",
                    full.names = T)
# Subset first 48 samples
files <- files[1:48]
# Apply function and use map_dfr to combine dataframes together
qscore_data <- map_dfr(files, extract_nano_qscores)
# Save result to a rds file
saveRDS(qscore_data, "data/nanopore_quality.rds")
```

Violin plot

```{r ont_quality_plot, size = "small", fig.dim = c(6.5,10)}
# Read data
qdata <- readRDS("data/nanopore_quality.rds")

# Divide samples to groups of 12
sample_list <- unique(qdata$sampleid)
sample_groups <- split(sample_list, ceiling(seq_along(sample_list) / 12))

# Generate a separate plot for each group
plots <- map(sample_groups, function(group_samples) {
  ggplot(filter(qdata, sampleid %in% group_samples),
         aes(x = sampleid, y = nanopore_qscore, fill = sampleid)) +
    geom_violin(scale = "width", alpha = 0.6) +
    theme_fivethirtyeight(base_size=8) +
    labs(x = "Sample", y = "Nanopore Q-score") +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
})

# Combine all plots into a single vertical layout
final_plot <- wrap_plots(plots, ncol = 1)

# Show the plot
print(final_plot)
```

#### Import sequence data to Qiime 2

The create_manifest.sh script inside root directory can be used for creation of manifest file needed to import demuxed and trimmed sequence files.

```{bash, eval = F, size = "small"}
# Activate qiime environment
export PATH="/homedir04/msuokas/miniconda3/bin:$PATH"
source activate qiime-2024.10

# Import sequence data
qiime tools import \
  --type 'SampleData[SequencesWithQuality]' \
  --input-path data/reads/manifest.csv \
  --output-path data/demux.qza \
  --input-format SingleEndFastqManifestPhred33
```

#### Dereplicate sequences

Removes sequence redundancy.

```{bash, eval = F, size = "small"}
# Activate qiime environment
export PATH="/homedir04/msuokas/miniconda3/bin:$PATH"
source activate qiime-2024.10

# Dereplicate sequences with vsearch plugin
qiime vsearch dereplicate-sequences \
    --o-dereplicated-table data/derep_table.qza \
    --o-dereplicated-sequences data/derep_seqs.qza \
    --i-sequences data/demux.qza
```

#### Pick de-novo otus

Clustering of sequences at 97 % identity level

```{bash, eval = F, size = "small"}

# Activate qiime environment
export PATH="/homedir04/msuokas/miniconda3/bin:$PATH"
source activate qiime-2024.10

# Cluster
qiime vsearch cluster-features-de-novo \
   --i-sequences data/derep_seqs.qza \
   --i-table data/derep_table.qza \
   --p-strand plus --p-threads 10 --p-perc-identity 0.97 \
   --o-clustered-table data/clustered-table.qza \
   --o-clustered-sequences data/clustered-seqs.qza

```

\newpage

#### Filter rare features

```{bash, eval = F, size = "small"}

# Activate qiime environment
export PATH="/homedir04/msuokas/miniconda3/bin:$PATH"
source activate qiime-2024.10

# Filter table
qiime feature-table filter-features  \
--i-table data/clustered-table.qza  \
--p-min-frequency 10 \
--o-filtered-table data/n10-table.qza

# Filter sequences
qiime feature-table filter-seqs \
--i-data data/clustered-seqs.qza \
--i-table data/n10-table.qza \
--o-filtered-data data/n10-seqs.qza
```

#### Detect and remove chimeric features

```{bash, eval = F, size = "small"}

# Activate qiime environment
export PATH="/homedir04/msuokas/miniconda3/bin:$PATH"
source activate qiime-2024.10

# Detect chimeric features
 qiime vsearch uchime-denovo \
 --i-table data/n10-table.qza \
 --i-sequences data/n10-seqs.qza \
 --o-nonchimeras data/nonchimeric.qza \
 --o-stats data/stats.qza \
 --o-chimeras data/chimeric.qza
 
 # Clean table
 qiime feature-table filter-features \
  --i-table data/n10-table.qza \
  --m-metadata-file data/nonchimeric.qza \
  --o-filtered-table data/feature-table.qza
  
  # Clean sequences
  qiime feature-table filter-seqs \
  --i-data data/n10-seqs.qza \
  --i-table data/feature-table.qza \
  --o-filtered-data data/feature-seqs.qza 
 
 
```

#### Import feature table and metadata to R

```{r import, size = "small"}
#sequence file
tse <- importQIIME2(featureTableFile = "data/feature-table.qza")
tse <- tse[, sort(colnames(tse))]
#add metadata
metadata <- data.frame(read_tsv("data/jt_meta.tsv",
                                show_col_types = F))
metadata <- column_to_rownames(metadata, "SampleID")
colData(tse) <- DataFrame(metadata)
tse
```

#### Assign taxonomy

Confirm correct sequence order first as QIIME shuffles natural order.

```{r wrangle_sequence_order, size="small"}
ref_sequences <- importQZA("data/feature-seqs.qza")
ref_ids <- names(ref_sequences)
tse_ids <- rownames(tse)
# Check if all rownames are present in the reference IDs
if (!all(tse_ids %in% ref_ids)) {
  stop("Not all rownames from tse are present in the reference sequences.")
}
# Reorder `ref_sequences` to match the order of `tse` rownames
ref_sequences_ordered <- ref_sequences[match(tse_ids, ref_ids)]
all(names(ref_sequences_ordered) == rownames(tse))
referenceSeq(tse) <- ref_sequences_ordered
tse
```

Assign taxonomy using Silva reference database.

```{r, size = "small", eval = F}
# assign taxa
taxa <- assignTaxonomy(referenceSeq(tse), minBoot=90, multithread=10,
        refFasta="~/reference/silva_nr99_v138.1_train_set.fa.gz")
# add species column (empty)
taxa <- data.frame(taxa)
taxa$species <- "NA"
saveRDS(taxa, "data/taxa.rds" )
```

Include taxonomy table

```{r, size = "small"}
# Read taxa file
taxa <- readRDS("data/taxa.rds")
# Add to rowData
rowData(tse) <- taxa
# Rename rows
rownames(tse) <- paste0("OTU", seq(1,nrow(tse)))
# Check tse
tse
```

#### Write object data to files

Write RDS. File can be reloaded as an unchanged object in R

```{r rds, size = "small"}
saveRDS(tse, "results/tse.rds")
```

Write abundance table

```{r otus, size = "small"}
# FeatureID will be rowname
abd <- data.frame(FeatureID = rownames(tse),assays(tse)$counts)
# Write
write_tsv(abd, "results/feature_table.tsv")
```

Write taxonomy table

```{r taxa, size = "small"}
# FeatureID will be rowname
taxt <- data.frame(FeatureID = rownames(tse), rowData(tse))
# Write
write_tsv(taxt, "results/taxonomy.tsv")
```

Feature sequences to fasta file

```{r write_seq, size = "small"}
writeXStringSet(referenceSeq(tse), "results/repseq.fasta",
                                      append = F, compress = F,
                                      format = "fasta")
```

\newpage

Write metadata file

```{r write_meta, size = "small"}
metadf <- data.frame(colData(tse)) %>% rownames_to_column(var="Sampleid")
# Write
write_tsv(metadf, "results/metadata.tsv")
```

#### Microbe data analysis

Agglomeration of taxonomy to genus rank and count relative abundance

```{r prepare, size = "small"}
# Agglomeration
altExp(tse, "Genus") <- agglomerateByRank(tse, rank = "Genus",
                             onRankOnly = T, na.rm = F)
# Calculate relabundance
altExp(tse, "Genus") <- transformAssay(altExp(tse, "Genus"),
                                       assay.type = "counts",
                                       method = "relabundance")
```

Pick ten most abundant features

```{r topten, size = "small"}
#top10 features
top10 <- getTop(altExp(tse, "Genus"), top = 10,
                        method = "mean",
                        assay.type = "relabundance")
#create and filter table
table <- data.frame(assays(altExp(tse, "Genus"))$relabundance)
table <- table %>% rownames_to_column(var = "Genus") %>% 
  filter(Genus %in% top10)
```

#### Composition plot

Change data to long table format

```{r barplot, size = "small", fig.dim = c(7,5)}
df_long <- table %>% pivot_longer(cols = starts_with("nr"),
names_to = "Sample", values_to = "Abundance")
#Plot stacked barplot
ggplot(df_long, aes(x = Sample, y = Abundance, fill = Genus)) +
geom_bar(stat = "identity") +
   theme_fivethirtyeight(base_size = 8) +
   scale_fill_brewer(palette = "Spectral") +
   theme(axis.text.x =  element_text(angle = 90))
```

#### Summary of otu picking strategy

Create summary table

```{r comparison, size = "small"}
summary <- data.frame(Sample = colnames(tse),
                      Counts = colSums(assays(tse)$counts),
                      Features = colSums(assays(tse)$counts != 0))
                      
rownames(summary) <- NULL
kable(summary, digits=0) %>% 
  kable_styling(latex_options = c("HOLD_position", "striped",
                                  "repeat_header"),
                font_size = 8) %>%
  row_spec(0, background = "teal", color = "ivory")
```

#### Alpha diversity

```{r shannon, size = "small"}
tse <- addAlpha(tse, assay.type="counts", index="shannon_diversity",
                sample=10000)
shannon <- data.frame(Samples = colnames(tse),
                      Shannon_index = colData(tse)$shannon_diversity)
rownames(shannon) <- NULL

# Table
kable(shannon, digits=2, caption = "Shannon index") %>%
kable_styling(latex_options = c("HOLD_position", "striped",
                                "repeat_header"),
              font_size = 8) %>% row_spec(0, background = "teal",
                                           color = "ivory")
```

\newpage

#### Beta diversity measured by Bray-Curtis dissimilarity

```{r bray, size = "small"}
tse <- runMDS(tse, FUN = getDissimilarity, method = "bray",
              assay.type = "counts", name = "MDS_bray", sample=10000)
#Explained variance
e <- attr(reducedDim(tse, "MDS_bray"), "eig")
rel_eig <- e / sum(e[e > 0])
#Plot
plotReducedDim(tse, "MDS_bray", colour_by = "Source",
               text_by = "Type", text_size = 3, shape_by = "Type") +
  labs(x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1),
                 "%", ")", sep = ""), y = paste("PCoA 2 (",
                 round(100 * rel_eig[[2]], 1), 
                 "%", ")", sep = "")) + scale_color_lancet()
```
